---
title: "R Conditionals and Iteration"
weight: 4
output:
  pdf_document: default
  html_document: default
categories: ["R Programming"]
---

<p class="lead">This workshop will explore approaches for implementing logic, testing conditions and iterating over R objects. The material will include comparisons via operators, looping, vectorization, "apply" functions and control flow. This lesson assumes a [basic familiarity with R](/lesson/r-intro).</p>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

The ability to use logic to condition execution of a program is fundamental to many languages. These *control-flow* features are primarily implemented in R through a series of "reserved" words:

```{r, eval = FALSE}
?reserved
```

We can access the documentation on R's implementation of control flow statements through any of the relevant reserved words:

```{r, eval = FALSE}
?`if`
?`else`
?`for`
?`break`
?`while`
?`repeat`
```

Note the back ticks around the names of these statements. The reservation on these names prevents them from being assigned another value. It also introduces a concept that is important, particularly for more advanced programming tasks in R: 

> *Everything that happens is a function call[^1].*

Although the syntax is a bit different, these statements are actually functions that can accept arguments and return values. As we saw in the control flow help file, the parameters can include:

- `cond` = a logical condition (i.e. `TRUE` or `FALSE`) ... used with `if` and `while`
- `var` = a placeholder name for one member of a sequence ... used with `for`
- `seq` = a sequence over which to iterate ... used with `for`
- `expr` = the "result" or expression to evaluate after logic is complete ... used with *most* statements

We can illustrate how these statements behave as functions with the following:

```{r, eval = FALSE}
`if`(cond = months(Sys.Date()) %in% month.name[6:8], 
     cons.expr = "summer!",
     alt.expr = "not summer!")
```

The syntax above is a bit arcane ... and we'll discuss a more common (and legible) way to formulate an if/else statement below. But the principle that these statements are functions is helpful in understanding how they behave.

### `if` / `else`

R provides some helpful syntax for using them to implement the functions for `if` and `else` conditional statements.

Let's return to the example above. 

The if statement could be written as we have it ... but more colloquially would be written as follows:

```{r}
if (months(Sys.Date()) %in% month.name[6:8]) {
  
  "summer!"
  
} else {
  
  "not summer!"
  
}
```

We can break down the logic by starting with some pseudo code of what we want to achieve:

```
if the current month is june, july or august
  print "summer!"
otherwise 
  print "not summer!"
```

This logic operates on the following question: 

Is the current month ... 

a. either June, July or August
b. neither June, July nor August

As we can see, the answer to this question is binary, and can therefore evaluate to a simple `TRUE` or `FALSE`.

We construct this test using `%in%`, which is a built-in operator for determining whether or not there are matching values between vectors. `months(Sys.Date())` gives us a character vector of length 1 with the full name of the current month.

```{r}
Sys.Date()
months(Sys.Date())
```

And `month.name[6:8]` gives us a vector of length 3 with the full names of the 6th, 7th and 8th months of the year.

```{r}
month.name
month.name[6:8]
```

So is the current month one of these three?

```{r}
months(Sys.Date()) %in% month.name[6:8]
```

We might consider saving ourselves a few keystrokes and assigning this test to an object:

```{r}
summer_test <- months(Sys.Date()) %in% month.name[6:8]
```

From there we can use the expressions (code inside of `{` squigly braces) following `if` and `else` to define what action we should take on `TRUE` and `FALSE` respectively. In this case it's simply a matter of printing different strings.

#### `else if`

`if` / `else` statements don't have to stop after one logical test. Given that the first `if` fails, you can continue to condition the statement arbitrarily with `else if`: 

```{r}
if (summer_test) {
  
  "summer!"
  
} else if (weekdays(Sys.Date()) == "Saturday") {
  
  "not summer ... but it's saturday!"
  
} else {
  
  "not summer. not saturday."
  
}
  
```

Note that the final `else` is very important, as it closes out the logical possibilities. Without it, the statement wouldn't print anything if both the `if` and `else if` fail.

As an aside, we can also make the logical argument to `if` / `else` more complex with boolean operators. For example if we wanted to satisfy two conditions simultaneously we could use `&&`:

```{r}
x <- 12

if (x > 10 && x%%2 == 0) {
  
  "big and even"
  
} else {
  
  "small or odd"
  
}

x <- 13

if (x > 10 && x%%2 == 0) {
  
  "big and even"
  
} else {
  
  "small or odd"
  
}
```

And if we wanted to test either one or both of the conditions simultaneously, we could use `||`:

```{r}
x <- 4

if (x > 10 || x%%2 == 0) {
  
  "big or even"
  
} else {
  
  "small or odd"
  
}

x <- 3

if (x > 10 && x%%2 == 0) {
  
  "big or even"
  
} else {
  
  "small or odd"
  
}
```

#### `ifelse()`

For brevity, there is a single function to manage `if` / `else` control flow. `ifelse()` takes three arguments:

- `test` = a condition that evalues to `TRUE` or `FALSE`
- `yes` = the value(s) to be returned for `TRUE`
- `no` = the value(s) to be returned for `FALSE`

`ifelse` can be particularly useful as a helper function. Here's an example of using it in a data manipulation pipeline with **dplyr**:

```{r, message=FALSE, warning=FALSE}
library(dplyr)

mtcars %>% 
  mutate(transmission = ifelse(am == 1, "Manual", "Automatic")) %>%
  select(transmission, mpg)
```

The function can behave very similarly to the traditional `if` / `else` statement:

```{r}
if (summer_test) {
  
  "summer!"
  
} else {
  
  "not summer!"
  
}
```

```{r}
ifelse(summer_test, "summer!", "not summer!")
```

Both of the cases above achieve the same result, but there are a couple of fundamental differences between the two approaches. `?ifelse` gives us some insight to the nature of the values the function returns: 

> A vector of the same length and attributes (including dimensions and "class") as test and data values from the values of yes or no. 

We can see that the return value is a *vector* ... `ifelse()` has a distinct advantage over the more explicit `if` / `else` syntax in that it is **vectorized**. We'll talk more about vectorization and its advantages in the context of iteration and loops, but we can see some of its benefits already with `ifelse()`.

For example, say we had a test that evaluated multiple elements:

```{r}
x <- 1:10

x > 5
```

If we were to use this in a traditonal `if` / `else` statement, we would get a warning:

```{r}
if (x > 5) {
  
  "large"
  
} else {
  
  "small"
  
}
```

But with `ifelse()`, we get a return vector of the same length as the test:

```{r}
ifelse(x > 5, "large", "small")
```

Note that this vectorization can result in some potentially unexpected behavior. As the examples in the help file for `ifelse()` indicate, the function can occassionally strip important attributes for the return vector. Let's try the following to demonstrate:

```{r}
ifelse(1 < 2, Sys.Date(), NA)
```

What's that all about? Weren't we returning a *date* for `TRUE`?

As it turns out, `ifelse` has stripped the date attribute:

```{r}
class(ifelse(1 < 2, Sys.Date(), NA))
```

And what we are left with is the numeric representation of the date (i.e. days since origin, which defaults to "1970-01-01"):

```{r}
Sys.Date()
as.numeric(Sys.Date())

as.Date(as.numeric(Sys.Date()), origin = "1970-01-01")

dput(as.Date(as.numeric(Sys.Date()), origin = "1970-01-01"))
```

The more explicit `if` / `else` syntax, on the other hand, will preserve the date attributes:

```{r}
if(1 < 2) {
  
  Sys.Date()
  
  } else {
    
    NA
    
    }
```

```{r}
class (
  
  if(1 < 2) {
    
    Sys.Date()
    
    } else {
      
      NA 
    }
  
)
```

#### `switch()`

`switch()` provides yet another way to evaluate conditional logic in R. This function can be helpful for saving time with coding (and making code more legibile), particularly when there are a number of conditions to satisfy[^2].

```{r}
switch(EXPR = "foo", 
       "foo" = 1:10,
       "bar" = 11:20,
       "baz" = 21:30,
       "bop" = 31:40)

switch(EXPR = "bar", 
       "foo" = 1:10,
       "bar" = 11:20,
       "baz" = 21:30,
       "bop" = 31:40)

switch(EXPR = "baz", 
       "foo" = 1:10,
       "bar" = 11:20,
       "baz" = 21:30,
       "bop" = 31:40)

switch(EXPR = "bop", 
       "foo" = 1:10,
       "bar" = 11:20,
       "baz" = 21:30,
       "bop" = 31:40)
```

### `for`

`for` allows us to iterate (or *loop*) over multiple elements. The general format for implementing a for loop in R is:

```{r, eval = FALSE}
for (i in vector) {
  
  # code here
  # usually doing something with vector[i]
  # where vector is an existing vector
  # and i is an arbitrary representation of a single element of that vector
  
}
```

Let's return to the example where we tried to use `if` / `else` with mulitple conditions:

```{r}
x <- 1:10

if (x > 5) {
  
  "large"
  
} else {
  
  "small"
  
}
```

We saw how to get around this warning by using the vectorized `ifelse()`. But we can actually use `if` / `else` as well, we just need to do so inside of a for loop:

```{r}
for (i in x) {
  
  if (i > 5) {
    
    print("large")
    
  } else {
    
    print("small")
    
  }
  
}
```

Note that in order to test each element in `x` we subset it at an index we are arbitrarily calling `i`, which represents each iteration. This is arbitrary in that we could have used another name (`var`, `num`, etc) for this index. But the vector `x` *does* need to exist in the environment. 

It's also worth pointing out that we've added an explicit `print()` around the return values. We need this for the output because automatic printing is turned off inside of loops[^3].

### `while`

`while` is another looping construct that you can use in R. The idea for while is to:

1. evaluate a logical condition
2. execute an action if that condition evaluates to `TRUE`
3. continue executing that action until that condition is `FALSE`
```{r, eval = FALSE}
st <- Sys.time()
et <- Sys.time()

while (et < st + 5) {
  print("busy")
}
```

If we run the code above we see an *infinite loop* of `"busy"` ... oops.

As described above, the `while` loop will continue executing as long as the condition remains `TRUE`. So in many cases you need to explicitly advance or modify what is being conditionally evaluated to avoid getting caught in a continuous loop.

```{r, eval = FALSE}
st <- Sys.time()
et <- Sys.time()

while (et < st + 5) {
  print("busy")
  et <- Sys.time()
}
```

### `repeat`

If we want a loop to re-use the same expression over and over again, we can construct a `repeat` loop:

```{r, eval = FALSE}
repeat {
  
  x <- sample(1:20,1)
  
  if(x == 2) {
    print("finally!")

  } else {
    
    print(x)
    
  }
  
}
```

The code above will repeat the same expression ... but as with our original `while` loop, it will do so infinitely. To make this kind a looping process effective, we need to be able to exit out of the loop 

### `break`

There are several ways to navigate inside of a loop:

- `break` = cancel execution and exit out of the loop silently
- `stop()` = cancel execution and error out of the loop
- `next` = proceed to the following iteration

Going back to our `repeat` loop, we can actually resolve the infinite nature of the loop with `break`:

```{r}
repeat {
  
  x <- sample(1:20,1)
  
  if(x == 2) {
    print("finally!")
    break

  } else {
    
    print(x)
    
  }
  
}
```

Let's set up another example, which we'll use to compare `break`, `stop()` and `next`:

```{r}
for (num in 2:20) {
  
  if (num %% 2 == 0) {
    
    print(num)
    
  } else {
    
    break
    
  }
  
}
```


### `stop()`

We can use `stop()` to cancel execution as well:

```{r, error = TRUE}
for (num in 2:20) {
  
  if (num %% 2 == 0) {
    
    print(num)
    
  } else {
    
    stop("not even")
    
  }
  
}
```

The important thing to point out with `stop()`, however, is that it behaves like a regular function (i.e. not a reserved word like `break`) and it returns an *error*.

### `next`

If we want advance through to the subsequent element rather than breaking or stopping execution altogether, we can use `next`:

```{r}
for (num in 2:20) {
  
  if (num %% 2 == 0) {
    
    print(num)
    
  } else {
    
    next
    
  }
  
}
```

### "apply" functions

As an alternative to looping (and in particular `for` loops), R provides *functionals* for iterating across multiple elements. These are typically referred to as the "apply" family of functions, and they generally allow you to pass arbirtrary functions to each item in a vector or list.

There are a number of different apply functions, each which has its own use-case depending on the type of object over which it will be applied and the type of value it will return. We'll look at three of the most commonly used:

- `apply()` = useful for applying functions across rows or columns of matrices or data frames
- `lapply()` = passes an abritrary function across a list or vector and returns a list
- `sapply()` = a "simplified" version of `lapply()` ... returns a vector

For `apply()` we define the matrix or data frame we are working with ("X"), the dimension across which we want to apply a function ("MARGIN"; 1 for rows, 2 for columns) and the function to use ("FUN"):

```{r}
apply(X = mtcars, MARGIN = 1, FUN = max)
```

```{r}
apply(X = mtcars, MARGIN = 2, FUN = mean)
```

`lapply()` ("list apply") operates in a similar way but doesn't need a specification for "MARGIN" since it works on lists and vectors:

```{r}
lapply(X = list(1:10, 5:20, 100:400), FUN = mean)
```

Note that this returns a list. 

```{r}
class(lapply(X = list(1:10, 5:20, 100:400), FUN = mean))
```

`sapply()` with the same arguments "simplifies" the return value to a vector:

```{r}
sapply(X = list(1:10, 5:20, 100:400), FUN = mean)
```

As we've seen `sapply()` and `lapply()` each work on lists. They also each work on vectors:

```{r}
lapply(c("dog", "cat", "rat"), grepl, "dog")
```

```{r}
sapply(c("dog", "cat", "rat"), grepl, "dog")
```

By default `sapply()` returns a named vector, whereas `lapply()` returns a list with no names. 

But it's worth noting that these are fundamentally doing the same thing, just with slightly different outputs by default. If we wanted to coerce `sapply()` to give us an identical result to `lapply()`, we could:

```{r}
lres <- lapply(c("dog", "cat", "rat"), grepl, "dog")
lres
```

```{r}
sres <- sapply(c("dog", "cat", "rat"), grepl, "dog", simplify = FALSE, USE.NAMES = FALSE)
sres
```

```{r}
identical(lres, sres)
```

Also keep in mind that with the apply family we can also use *anonymous* functions, which are those that we write "on-the-fly" and are only available within the environment of the function in which they are executed:

```{r}
sapply(X = 1:10, FUN = function(x) x%%2 == 0)
```

These apply functions can in some cases enhance performance when iterating over large objects. But a much more efficient approach to programming with many iterations is vectorization.

### Vectorization

As we've seen, looping can help us implement some advanced logic in our code. But can introduce challenges in terms of efficiency. Here we'll briefly touch on alternative, *vectorized* approaches.

Let's set up a simple loop:

```{r}
x <- vector()
  
for (i in 1:1e7) {
  
  x[i] <- i + i^2
    
}
```

That seemed to take a while ... let's time it:

```{r}

system.time({
  
  x <- vector()
  
  for (i in 1:1e7) {
    
    x[i] <- i + i^2
    
  }
  
})
```

Let's try the same thing with `sapply()`:

```{r}
y <- sapply(1:1e7, function(x) x + x^2)
```

That also took a while. How long?

```{r}
system.time({
  
  y <- sapply(1:1e7, function(x) x + x^2)

})
```

OK ... so the `for` loop and `apply` function were both about the same speed. What if we try a *vectorized* approach:

```{r}
z <- 1:1e7 + (1:1e7)^2
```

Much faster.

```{r}
system.time({
  
  z <- 1:1e7 + (1:1e7)^2

})
```

```{r}
identical(x,y,z)
```

Vectorization can significantly increase the speed of code execution. The function being executed at each element of the vector *knows* the data type of the given element (because it is a homogenous vector), and can therefore skip any overhead involved with typing[^4][^5].

Futhermore, the kinds of `for` loops we've written above force R to first create a vector ... then retrieve it from memory ... then add a new element ... and repeat for the length of the vector over which we are iterating.

In some cases you can significantly speed up a loop by predefining a vector *with* the appropriate type (passed in the *mode* argument) and length:

```{r}
x <- vector(mode = "numeric", length=1e7)
  
for (i in 1:1e7) {
  
  x[i] <- i + i^2
    
}
```

```{r}
system.time({
  
  x <- vector(mode = "numeric", length=1e7)
    
  for (i in 1:1e7) {
    
    x[i] <- i + i^2
      
  }
  
})
```

### Further Resources

- http://adv-r.had.co.nz/
- https://www.datamentor.io/r-programming/
- https://www.tutorialgateway.org/r-programming/

[^1]: http://user2014.r-project.org/files/chambers.pdf]
[^2]: https://stackoverflow.com/questions/7825501/switch-statement-usage 
[^3]: https://stackoverflow.com/questions/4716152/why-do-r-objects-not-print-in-a-function-or-a-for-loop?answertab=votes#tab-top
[^4]: http://www.noamross.net/blog/2014/4/16/vectorization-in-r--why.html
[^5]: http://alyssafrazee.com/2014/01/29/vectorization.html


```{r, include=FALSE, eval = TRUE}
# find packages attached during for this Rmd session and their versions
package <- installed.packages()[names(sessionInfo()$otherPkgs), "Package"]
version <- installed.packages()[names(sessionInfo()$otherPkgs), "Version"]

thesepkgs <- data.frame(package, version)

# if no non-base packages installed skip
if(nrow(thesepkgs) == 0) invisible()

# find package csv file
fp <- "../../packages.csv"

# if it exists find read in contents ... combine with attached pkgs and dedupe
if(file.exists(fp)) {
  
  pkgs <- read.csv(fp)
  
  pkgs <- 
    rbind(pkgs, thesepkgs)
  
  pkgs <- pkgs[!duplicated(pkgs),]
  
} else{
  
  pkgs <- thesepkgs
  
}

# write out new package.csv file
write.table(pkgs, 
          file = fp,
          sep = ",",
          row.names = FALSE)
```



